---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R package: covidprobability

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

This package provides the functions, data and documentation that support a 
calculator to determine the probability of an undetected COVID-19 infection in
a setting/unit after a potential exposure, testing, and when there are no 
symptomatic cases. For a detailed explanation of the rationale and 
implementation, please see the [vignette](https://eebrown.github.io/covidprobability/articles/unit-example.html).

## Shiny

An [interactive web app](http://covid.eebc.ca) of this calculator is available.

## Installation

You can install the latest version of covidprobability from [Github](https://github.com/eebrown/covidprobability) with:

``` r
devtools::install.github("eebrown/covidprobability")
```

### Disclaimer

This is an exploratory model and may contain errors. Please see the [vignette](https://eebrown.github.io/covidprobability/articles/unit-example.html) for assumptions and limitations of the model. It should not be relied upon for clinical decisions.

## Example


```{r example}

library(covidprobability)

test_n <- unit_probability(test_day = 9, pre0 =  0.13, sens = sens, spec = 1, 
                           asympt = 0.279, days = 0:14, mu = 1.63, sigma = 0.5, 
                           n = 10)

```


```{r unit_example, echo = FALSE}

plot(1:14, test_n$point, type="b", ylim=c(0,1),
     main = "Unit-wide probabiltiy of undetected COVID-19",
     xlab = "Days since exposures",
     ylab = "Probability")
lines(test_n$lower, type="l", col="grey")
lines(test_n$upper, type="l", col="grey")
abline(v = 9)
text(1:14, (test_n$point + 0.1), round(test_n$point, 2), cex = 1)

```



